Simple repression draft

### 2020-06-03 doc/appendix_01_nonbursty.tex
MRM quick pass through appendix 1. looks really good

@@

Next consider the cartoon for nonequilibrium model two in
Figure~\ref{fig1:means_cartoons}. Now we must track probabilities
$p_R$, $p_P$, and $p_E$ for the repressor bound, empty, and
polymerase bound states, respectively. By analogy
to~\ref{eq:poisson_promoter_cme_appdx}, the master equation for
model two can be written
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_E(m,t) =&\; 
  \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_E(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_E(m,t)}^{m \rightarrow m - 1}.
\\
&
+ \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
- \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{rp_P(m-1,t)}^{m-1 \rightarrow m,\;A \rightarrow U}
\\
\deriv{t}p_P(m,t) =&\; 
- \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
+ \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{(m+1)\gamma p_P(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_P(m,t)}^{m \rightarrow m - 1}.
\\
&- \overbrace{rp_P(m,t)}^{m \rightarrow m + 1,\;A \rightarrow U}.
\label{eq:model2_cme_appdx}
\end{split}
\end{align}
$k_P^+$ and $k_P^-$ are defined in close analogy to $k_R^+$ and
$k_R^-$, except for RNAP binding and unbinding instead of
repressor. Similarly $p_P(m,t)$ is defined for the active
(RNAP-bound) state exactly as are $p_R(m,t)$ and $p_E(m,t)$ for
the repressor bound and unbound states, respectively.
The new subtlety Eq.~\ref{eq:model2_cme_appdx} introduces compared to
Eq.~\ref{eq:poisson_promoter_cme_appdx} is that when mRNAs are
produced, the promoter state also changes. This is encoded by the
terms involving $r$, the last term in each of the equations for
$p_E$ and $p_P$. The former accounts for arrivals in the
unbound state and the latter accounts for departures from the
RNAP-bound state.

To condense and clarify the unwieldy notation of
Eq.~\ref{eq:model2_cme_appdx}, it can be rewriten in matrix form. We define the column vector $\vec{p}(m,t)$ as
\begin{equation}
\vec{p}(m,t)
= \begin{pmatrix} p_R(m,t) \\ p_E(m,t) \\ p_P(m,t) \end{pmatrix}
\end{equation}
to gather, for a given $m$, the probabilities of finding the
system in the three possible promoter states. Then all the
transition rates may be condensed into matrices which multiply
this vector. The first matrix is
\begin{equation}
\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
                        0 & k_P^+ & -k_P^- 
                \end{pmatrix},
% \label{eq:3state_cme_matrices_pt1}
\end{equation}
which tracks all promoter state changes in
Eq.~\ref{eq:model2_cme_appdx} that are \textit{not} accompanied
by a change in the mRNA copy number. The two terms accounting for
transcription, the only transition that increases mRNA copy
number, must be handled by two separate matrices given by
\begin{equation}
\mathbf{R_A} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & r \\ 
                0 & 0 & 0
                \end{pmatrix},\
\mathbf{R_D} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & 0 \\ 
                0 & 0 & r
                \end{pmatrix}.
% \label{eq:3state_cme_matrices_pt2}
\end{equation}
$\mathbf{R_A}$ accounts for transitions \textit{arriving} in a given state
while $\mathbf{R_D}$ tracks \textit{departing} transitions.
With these definitions, we can condense Eq.~\ref{eq:model2_cme_appdx} into
the single equation
\begin{equation}
\frac{d}{dt} \vec{p}(m,t) =
\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
                + \mathbf{R_A} \vec{p}(m-1,t) +
                \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
\label{eq:generic_cme_appdx}
\end{equation}
which is just Eq.~\ref{eq:3state_rep_cme} in the main text.
Straightforward if tedious algebra verifies that
Eqs.~\ref{eq:model2_cme_appdx} and~\ref{eq:generic_cme_appdx} are
in fact equivalent.

Although we derived Eq.~\ref{eq:generic_cme_appdx} for the particular case
of nonequilibrium model two in Figure~\ref{fig1:means_cartoons},
in fact the chemical master equations for all of the
nonequilibrium models in Figure~\ref{fig1:means_cartoons} except
for model five can be cast in this form. (We treat model five
separately in Section~\ref{sec:gen_fcn_appdx}.)
Model three introduces no new subtleties beyond model two and
Eq.~\ref{eq:generic_cme_appdx} applies equally well to it, simply
with different matrices of dimension four instead of three.
Models one and four are both handled by Eq.~\ref{eq:2state_rep_cme}
in the main text, which is just Eq.~\ref{eq:generic_cme_appdx}
except in the special case of $\mathbf{R_D} = \mathbf{R_A} \equiv
\mathbf{R}$, since in these two models transcription initiation
events do not change promoter state.

@@

\subsection{General forms for mean mRNA and Fano factor}
Our task now is to derive expressions for the first two moments of the
steady-state mRNA distribution from Eq.~\ref{eq:generic_cme_appdx}.
Our treatment of this is analogous to that given in~\cite{Sanchez2011}
and~\cite{Razo-Mejia2020}. We first obtain the steady-state limit of
Eq.~\ref{eq:generic_cme_appdx} in which the time derivative vanishes, giving
\begin{equation}
0 =
\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m)
                + \mathbf{R_A} \vec{p}(m-1) +
                \gamma (m+1) \mathbf{I} \vec{p}(m+1),
\label{eq:generic_cme_ss}
\end{equation}
From this, we want to compute
\begin{equation}
\langle{m}\rangle = \sum_S \sum_{m=0}^\infty m p_S(m)
\end{equation}
and
\begin{equation}
\langle{m^2}\rangle = \sum_S \sum_{m=0}^\infty m^2 p_S(m)
\end{equation}
which define the average values of $m$ and $m^2$ at steady state, where the
averaging is over all possible mRNA copy numbers and promoter states $S$.
For example, for model one in Figure~\ref{fig1:means_cartoons}, the sum on $S$
would cover repressor bound and unbound states ($R$ and $U$ respectively),
for model two, the sum would cover repressor bound, polymerase bound,
and empty states ($R$, $P$, and $E$), and so on for the other models.

@@

Next we can find the mean mRNA copy number $\langle{m}\rangle$ from
$\langle\vec{m}\rangle$ according to
\begin{equation}
\langle{m}\rangle = \vec{1}^\dagger\cdot\langle\vec{m}\rangle,
\label{eq:m_from_vecm}
\end{equation}
where $\vec{1}^\dagger$ is a row vector whose elements are all 1.
Eq.~\ref{eq:m_from_vecm} holds since the $i$-th element of the column vector
$\langle\vec{m}\rangle$ is the mean mRNA value conditional on the system
occupying the $i-$th promoter state, so the dot product with $\vec{1}^\dagger$
amounts to simply summing the elements of $\langle\vec{m}\rangle$.
Rather than solving Eq.~\ref{eq:generic_cme_deriv_040} for $\langle\vec{m}\rangle$
and then computing $\vec{1}^\dagger\cdot\langle\vec{m}\rangle$, we may
take a shortcut by muliplying Eq.~\ref{eq:generic_cme_deriv_040} by
$\vec{1}^\dagger$ first. The key observation that makes this useful is that
\begin{equation}
\vec{1}^\dagger \cdot (\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) = 0.
\end{equation}
Intuitively, this equality holds because each column of this matrix
represents transitions to and from a given promoter state. In any given column,
the diagonal encodes all departing transitions and off-diagonals encode all
arriving transitions. Conservation of probability means that each column
sums to zero, and summing columns is exactly the operation that multiplying by
$\vec{1}^\dagger$ carries out.

@@

Also note that we are computing Fano factors to verify the results of
Section~\ref{sec:beyond_means}, concerning the constitutive promoter models
in Figure~\ref{fig2:constit_cartoons} which are analogs of the simple
repression models in Figure~\ref{fig1:means_cartoons}. We can translate
the matrices from the simple repression models to the constitutive case
by simply all occurences of repressor rates and removing the row and column
corresponding to the repressor bound state. The results for
$\langle{m}\rangle$ computed in the repressed case can be easily translated
to the constitutive case, rather than recalculating from scratch, by taking
the limit $k_R^+\rightarrow 0$, since this amounts to sending repressor
copy number to zero.

@@

\subsubsection{Fano factor}
To verify that the Fano factor for model one in
Figure~\ref{fig2:constit_cartoons} is in fact 1 as claimed in the main text,
note that in this limit $p_U = 1$ and $\langle{m}\rangle = r/\gamma$. All
elements of $\mathbf{K}$ are zero, and $\mathbf{R_A}-\mathbf{R_D} = 0$,
so Eq.~\ref{eq:vecm_appdx_ref} reduces to
\begin{equation}
- \gamma \langle\vec{m}\rangle = - r,
\end{equation}
or, in other words, since there is only one promoter state,
$\langle\vec{m}\rangle = \langle{m}\rangle$. Then it follows that
\begin{equation}
\nu = 1 -\frac{r}{\gamma}
    + \frac{r \langle{m}\rangle}{\gamma \langle{m}\rangle}
= 1
\end{equation}
as claimed.

@@

\subsection{Nonequilibrium Model Three
    - Multistep Transcription Initiation and Escape}
\subsubsection{Mean mRNA}
In close analogy to model two above, nonequilibrium model three from
Figure~\ref{fig1:means_cartoons} can be described by our generic master equation
Eq.~\ref{eq:generic_cme_appdx} with promoter transition matrix given by
\begin{equation}
\mathbf{K} =
\begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
        0 & k_P^+ & -k_P^- - k_O & 0 \\
        0 & 0 & k_O & 0
\end{pmatrix}
\end{equation}
and transcription matrices given by
\begin{equation}
\mathbf{R_A} =
\begin{pmatrix}
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & r \\ 
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & 0
\end{pmatrix},\
\mathbf{R_D} =
\begin{pmatrix}
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & r
\end{pmatrix}.
\end{equation}
$\langle\vec{m}^0\rangle$ is again given by Eq.~\ref{eq:vecm0_appdx_ref},
which in this case takes the form
\begin{equation}
(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) \langle\vec{m}^0\rangle =
\begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
    k_R^- & -k_R^+ -k_P^+ & k_P^- & r \\
    0 & k_P^+ & -k_P^- - k_O & 0 \\
    0 & 0 & k_O & - r
\end{pmatrix}
\begin{pmatrix} p_R \\ p_E \\ p_C \\ p_O
\end{pmatrix} = 0,
\end{equation}
where the four components of $\langle\vec{m}^0\rangle$ correspond to the
four promoter states repressor bound, empty, RNAP-bound closed complex, and
RNAP-bound open complex. As explained in Section~\ref{sec:m0_appdx}, we
are free to discard one linearly dependent row from this matrix and replace
it with the normalization condition $p_R + p_E + p_C + p_O = 1$.
Using normalization to eliminate $p_R$ from the first row gives
\begin{equation}
p_E = (1 - p_C - p_O)\frac{k_R^-}{k_R^- + k_R^+}.
\end{equation}
If we substitute this in the third row, then the last two rows constitute
two equations in $p_C$ and $p_O$ given by
\begin{align}
k_P^+k_R^-(1-p_C-p_O) - (k_P^- + k_O)(k_R^+ + k_R^-) p_C &= 0
\\
k_O p_C - r p_O &= 0.
\end{align}
Solving for $p_C = p_O r/k_O$ in the second and substituting into the first
gives us our desired single equation in the single variable $p_O$, which is
\begin{equation}
k_P^+k_R^- - k_P^+k_R^-\left(1 + \frac{r}{k_O}\right)p_O
            - (k_P^- + k_O)(k_R^+ + k_R^-) \frac{r}{k_O}p_O = 0,
\end{equation}
and solving for $p_O$ we find
\begin{equation}
p_O = \frac{k_P^+ k_R^- k_O}{k_P^+ k_R^- k_O + r k_P^+ k_R^- +
                            r (k_P^- + k_O) (k_R^+ + k_R^-)}.
\label{eq:model3_pO}
\end{equation}
Once again $p_O$, the transcriptionally active state, is the only component
of $\langle\vec{m}^0\rangle$ that survives multiplication by $\mathbf{R_A}$,
and $\mathbf{R_A}\langle\vec{m}^0\rangle = r p_O$. So
\begin{equation}
\langle{m}\rangle =
    \frac{1}{\gamma}\vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}^0\rangle
= \frac{r}{\gamma}
    \frac{k_P^+ k_R^- k_O}{k_P^+ k_R^- k_O + r k_P^+ k_R^- +
                            r (k_P^- + k_O) (k_R^+ + k_R^-)},
\end{equation}
which equals Eq.~\ref{eq:model3_mean_m} in the main text.

>>>>

\mrm{I personally think that the paper so far has done a great job at being very 
pedagogical. That is why I am of the opinion that before jumping into the 
derivation of moments and so on there needs to be a $p(m, t + \Delta t)$ kind of 
derivation.}

@@

Next consider the cartoon for nonequilibrium model two in
Figure~\ref{fig1:means_cartoons}(C). Now we must track probabilities $p_R$,
$p_P$, and $p_E$ for the repressor bound, empty, and polymerase bound states,
respectively. By analogy to~\ref{eq:poisson_promoter_cme_appdx}, the master
equation for model two can be written
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_E(m,t) =&\; 
  \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_E(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_E(m,t)}^{m \rightarrow m - 1}.
\\
&
+ \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
- \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{rp_P(m-1,t)}^{m-1 \rightarrow m,\;A \rightarrow U}
\\
\deriv{t}p_P(m,t) =&\; 
- \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
+ \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{(m+1)\gamma p_P(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_P(m,t)}^{m \rightarrow m - 1}.
\\
&- \overbrace{rp_P(m,t)}^{m \rightarrow m + 1,\;A \rightarrow U}.
\label{eq:model2_cme_appdx}
\end{split}
\end{align}
$k_P^+$ and $k_P^-$ are defined in close analogy to $k_R^+$ and $k_R^-$, except
for RNAP binding and unbinding instead of repressor. Similarly $p_P(m,t)$ is
defined for the active (RNAP-bound) state exactly as are $p_R(m,t)$ and
$p_E(m,t)$ for the repressor bound and unbound states, respectively. The new
subtlety Eq.~\ref{eq:model2_cme_appdx} introduces compared to
Eq.~\ref{eq:poisson_promoter_cme_appdx} is that when mRNAs are produced, the
promoter state also changes. This is encoded by the terms involving $r$, the
last term in each of the equations for $p_E$ and $p_P$. The former accounts for
arrivals in the unbound state and the latter accounts for departures from the
RNAP-bound state.

To condense and clarify the unwieldy notation of Eq.~\ref{eq:model2_cme_appdx},
it can be rewritten in matrix form. We define the column vector $\vec{p}(m,t)$
as
\begin{equation}
\vec{p}(m,t)
= \begin{pmatrix} p_R(m,t) \\ p_E(m,t) \\ p_P(m,t) \end{pmatrix}
\end{equation}
to gather, for a given $m$, the probabilities of finding the system in the three
possible promoter states. Then all the transition rates may be condensed into
matrices which multiply this vector. The first matrix is
\begin{equation}
\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
                        0 & k_P^+ & -k_P^- 
                \end{pmatrix},
% \label{eq:3state_cme_matrices_pt1}
\end{equation}
which tracks all promoter state changes in Eq.~\ref{eq:model2_cme_appdx} that
are \textit{not} accompanied by a change in the mRNA copy number. The two terms
accounting for transcription, the only transition that increases mRNA copy
number, must be handled by two separate matrices given by
\begin{equation}
\mathbf{R_A} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & r \\ 
                0 & 0 & 0
                \end{pmatrix},\
\mathbf{R_D} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & 0 \\ 
                0 & 0 & r
                \end{pmatrix}.
% \label{eq:3state_cme_matrices_pt2}
\end{equation}
$\mathbf{R_A}$ accounts for transitions \textit{arriving} in a given state while
$\mathbf{R_D}$ tracks \textit{departing} transitions. With these definitions, we
can condense Eq.~\ref{eq:model2_cme_appdx} into the single equation
\begin{equation}
\frac{d}{dt} \vec{p}(m,t) =
\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
                + \mathbf{R_A} \vec{p}(m-1,t) +
                \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
\label{eq:generic_cme_appdx}
\end{equation}
which is just Eq.~\ref{eq:3state_rep_cme} in the main text. Straightforward if
tedious algebra verifies that Eqs.~\ref{eq:model2_cme_appdx}
and~\ref{eq:generic_cme_appdx} are in fact equivalent.

Although we derived Eq.~\ref{eq:generic_cme_appdx} for the particular case of
nonequilibrium model two in Figure~\ref{fig1:means_cartoons}, in fact the
chemical master equations for all of the nonequilibrium models in
Figure~\ref{fig1:means_cartoons} except for model five can be cast in this form.
(We treat model five separately in Appendix~\ref{sec:gen_fcn_appdx}.) Model
three introduces no new subtleties beyond model two and
Eq.~\ref{eq:generic_cme_appdx} applies equally well to it, simply with different
matrices of dimension four instead of three. Models one and four are both
handled by Eq.~\ref{eq:2state_rep_cme} in the main text, which is just
Eq.~\ref{eq:generic_cme_appdx} except in the special case of $\mathbf{R_D} =
\mathbf{R_A} \equiv \mathbf{R}$, since in these two models transcription
initiation events do not change promoter state.

@@

\subsection{General forms for mean mRNA and Fano factor}
Our task now is to derive expressions for the first two moments of the
steady-state mRNA distribution from Eq.~\ref{eq:generic_cme_appdx}. Our
treatment of this is analogous to that given in~\cite{Sanchez2011}
and~\cite{Razo-Mejia2020}. We first obtain the steady-state limit of
Eq.~\ref{eq:generic_cme_appdx} in which the time derivative vanishes, giving
\begin{equation}
0 =
\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m)
                + \mathbf{R_A} \vec{p}(m-1) +
                \gamma (m+1) \mathbf{I} \vec{p}(m+1),
\label{eq:generic_cme_ss}
\end{equation}
From this, we want to compute
\begin{equation}
\langle{m}\rangle = \sum_S \sum_{m=0}^\infty m \, p_S(m)
\end{equation}
and
\begin{equation}
\langle{m^2}\rangle = \sum_S \sum_{m=0}^\infty m^2 p_S(m)
\end{equation}
which define the average values of $m$ and $m^2$ at steady state, where the
averaging is over all possible mRNA copy numbers and promoter states $S$. For
example, for model one in Figure~\ref{fig1:means_cartoons}(C), the sum on $S$
would cover repressor bound and unbound states ($R$ and $U$ respectively), for
model two, the sum would cover repressor bound, polymerase bound, and empty
states ($R$, $P$, and $E$), and so on for the other models.

@@

Next we can find the mean mRNA copy number $\langle{m}\rangle$ from
$\langle\vec{m}\rangle$ according to
\begin{equation}
\langle{m}\rangle = \vec{1}^\dagger\cdot\langle\vec{m}\rangle,
\label{eq:m_from_vecm}
\end{equation}
where $\vec{1}^\dagger$ is a row vector whose elements are all 1.
Eq.~\ref{eq:m_from_vecm} holds since the $i$-th element of the column vector
$\langle\vec{m}\rangle$ is the mean mRNA value conditional on the system
occupying the $i-$th promoter state, so the dot product with $\vec{1}^\dagger$
amounts to simply summing the elements of $\langle\vec{m}\rangle$. Rather than
solving Eq.~\ref{eq:generic_cme_deriv_040} for $\langle\vec{m}\rangle$ and then
computing $\vec{1}^\dagger\cdot\langle\vec{m}\rangle$, we may take a shortcut by
multiplying Eq.~\ref{eq:generic_cme_deriv_040} by $\vec{1}^\dagger$ first. The
key observation that makes this useful is that
\begin{equation}
\vec{1}^\dagger \cdot (\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) = 0.
\end{equation}
Intuitively, this equality holds because each column of this matrix
represents transitions to and from a given promoter state. In any given column,
the diagonal encodes all departing transitions and off-diagonals encode all
arriving transitions. Conservation of probability means that each column
sums to zero, and summing columns is exactly the operation that multiplying by
$\vec{1}^\dagger$ carries out.

@@

Also note that we are computing Fano factors to verify the results of
Section~\ref{sec:beyond_means}, concerning the constitutive promoter models in
Figure~\ref{fig2:constit_cartoons} which are analogs of the simple repression
models in Figure~\ref{fig1:means_cartoons}. We can translate the matrices from
the simple repression models to the constitutive case by simply substituting all
occurrences of repressor rates by zero and removing the row and column
corresponding to the repressor bound state. The results for $\langle{m}\rangle$
computed in the repressed case can be easily translated to the constitutive
case, rather than recalculating from scratch, by taking the limit
$k_R^+\rightarrow 0$, since this amounts to sending repressor copy number to
zero.

@@

\subsubsection{Fano factor}
To verify that the Fano factor for model one in
Figure~\ref{fig2:constit_cartoons}(A) is in fact 1 as claimed in the main text,
note that in this limit $p_U = 1$ and $\langle{m}\rangle = r/\gamma$. All
elements of $\mathbf{K}$ are zero, and $\mathbf{R_A}-\mathbf{R_D} = 0$, so
Eq.~\ref{eq:vecm_appdx_ref} reduces to
\begin{equation}
- \gamma \langle\vec{m}\rangle = - r,
\end{equation}
or, in other words, since there is only one promoter state,
$\langle\vec{m}\rangle = \langle{m}\rangle$. Then it follows that
\begin{equation}
\nu = 1 -\frac{r}{\gamma}
    + \frac{r \langle{m}\rangle}{\gamma \langle{m}\rangle}
= 1
\end{equation}
as claimed.

@@

\subsection{Nonequilibrium Model Three - Multistep Transcription Initiation and
Escape}

\subsubsection{Mean mRNA}
In close analogy to model two above, nonequilibrium model three from
Figure~\ref{fig1:means_cartoons}(C) can be described by our generic master
equation Eq.~\ref{eq:generic_cme_appdx} with promoter transition matrix given by
\begin{equation}
\mathbf{K} =
\begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
        0 & k_P^+ & -k_P^- - k_O & 0 \\
        0 & 0 & k_O & 0
\end{pmatrix}
\end{equation}
and transcription matrices given by
\begin{equation}
\mathbf{R_A} =
\begin{pmatrix}
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & r \\ 
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & 0
\end{pmatrix},\
\mathbf{R_D} =
\begin{pmatrix}
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & r
\end{pmatrix}.
\end{equation}
$\langle\vec{m}^0\rangle$ is again given by Eq.~\ref{eq:vecm0_appdx_ref},
which in this case takes the form
\begin{equation}
(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) \langle\vec{m}^0\rangle =
\begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
    k_R^- & -k_R^+ -k_P^+ & k_P^- & r \\
    0 & k_P^+ & -k_P^- - k_O & 0 \\
    0 & 0 & k_O & - r
\end{pmatrix}
\begin{pmatrix} p_R \\ p_E \\ p_C \\ p_O
\end{pmatrix} = 0,
\end{equation}
where the four components of $\langle\vec{m}^0\rangle$ correspond to the four
promoter states repressor bound, empty, RNAP-bound closed complex, and
RNAP-bound open complex. As explained in Section~\ref{sec:m0_appdx}, we are free
to discard one linearly dependent row from this matrix and replace it with the
normalization condition $p_R + p_E + p_C + p_O = 1$. Using normalization to
eliminate $p_R$ from the first row gives
\begin{equation}
p_E = (1 - p_C - p_O)\frac{k_R^-}{k_R^- + k_R^+}.
\end{equation}
If we substitute this in the third row, then the last two rows constitute two
equations in $p_C$ and $p_O$ given by
\begin{align}
k_P^+k_R^-(1-p_C-p_O) - (k_P^- + k_O)(k_R^+ + k_R^-) p_C &= 0
\\
k_O p_C - r p_O &= 0.
\end{align}
Solving for $p_C = p_O r/k_O$ in the second and substituting into the first
gives us our desired single equation in the single variable $p_O$, which is
\begin{equation}
k_P^+k_R^- - k_P^+k_R^-\left(1 + \frac{r}{k_O}\right)p_O
            - (k_P^- + k_O)(k_R^+ + k_R^-) \frac{r}{k_O}p_O = 0,
\end{equation}
and solving for $p_O$ we find
\begin{equation}
p_O = \frac{k_P^+ k_R^- k_O}{k_P^+ k_R^- k_O + r k_P^+ k_R^- +
                            r (k_P^- + k_O) (k_R^+ + k_R^-)}.
\label{eq:model3_pO}
\end{equation}
Once again $p_O$, the transcriptionally active state, is the only component of
$\langle\vec{m}^0\rangle$ that survives multiplication by $\mathbf{R_A}$, and
$\mathbf{R_A}\langle\vec{m}^0\rangle = r p_O$. So
\begin{equation}
\langle{m}\rangle =
    \frac{1}{\gamma}\vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}^0\rangle
= \frac{r}{\gamma}
    \frac{k_P^+ k_R^- k_O}{k_P^+ k_R^- k_O + r k_P^+ k_R^- +
                            r (k_P^- + k_O) (k_R^+ + k_R^-)},
\end{equation}
which equals Eq.~\ref{eq:model3_mean_m} in the main text.

\subsubsection{Fano factor}
To compute the Fano factor of the analogous constitutive promoter, we first
excise the repressor states and rates from the problem. More precisely, we
construct the matrix $(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A} -
\gamma\mathbf{I})$ and substitute it into Eq.~\ref{eq:vecm_appdx_ref} which is
now
\begin{equation}
(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A} - \gamma\mathbf{I})
    \langle\vec{m}\rangle
= \begin{pmatrix}
    -k_P^+ - \gamma & k_P^- & r \\
     k_P^+ & -k_P^- - k_O - \gamma & 0 \\
     0 & k_O & - r- \gamma
\end{pmatrix}
\begin{pmatrix} m_E \\ m_C \\ m_O \end{pmatrix}
= -r p_O \begin{pmatrix}1 \\ 0 \\ 0 \end{pmatrix}
\end{equation}
where we labeled the unbound, closed complex, and open complex components of
$\langle\vec{m}\rangle$ as $m_E$, $m_C$, and $m_O$, respectively. $p_O$ is given
by the limit of Eq.~\ref{eq:model3_pO} as $k_R^+\rightarrow 0$, which is
\begin{equation}
p_O = \frac{k_P^+  k_O}{k_P^+ (k_O + r) + r (k_P^- + k_O)}
\equiv \frac{k_P^+  k_O}{\mathcal{Z}},
\end{equation}
where we define $\mathcal{Z}$ for upcoming convenience as this sum of terms will
appear repeatedly. We can use the second equation to eliminate $m_E$, finding
$m_E = m_C(k_P^- + k_O + \gamma)/k_P^+$, and the third to eliminate $m_C$, which
is simply $m_C = m_O(r+\gamma)/k_O$. Substituting these both into the first
equation gives a single equation for the variable of interest, $m_O$,
\begin{equation}
-(k_P^+ + \gamma) (k_P^- + k_O + \gamma) (r + \gamma) m_O
    + k_P^- k_P^+ (r + \gamma) m_O + r k_P^+ k_O m_O = - r k_P^+ k_O p_O,
\end{equation}
which is solved for $m_O$ to give
\begin{equation}
m_O = p_O \frac{r k_P^+ k_O}
    {(k_P^+ + \gamma) (k_P^- + k_O + \gamma) (r + \gamma)
        -r k_P^+ k_O - k_P^- k_P^+ (r + \gamma)}.
\end{equation}
Expanding the denominator and canceling terms leads to
\begin{equation}
m_O = p_O \frac{r}{\gamma} \frac{k_P^+ k_O}
    {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2}.
\end{equation}
Now $\vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}\rangle = r m_O$, and
$\langle{m}\rangle = rp_O/\gamma$, so if we substitute these two quantities into
Eq.~\ref{eq:fano_appdx_ref}, we will readily obtain the Fano factor as
\begin{equation}
\nu = 1 - \langle{m}\rangle
    + \frac{\vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}\rangle}
            {\gamma \langle{m}\rangle}
= 1 - \frac{r}{\gamma}p_O + \frac{m_O}{p_O}.
\end{equation}
Substituting, we see that
\begin{equation}
\nu = 1 - \frac{r}{\gamma} \frac{k_P^+ k_O}{\mathcal{Z}}
    + \frac{r}{\gamma}
    \frac{k_P^+ k_O}
            {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2},
\end{equation}
and after simplifying, we obtain
\begin{equation}
\nu = 1 - \frac{r k_P^+ k_O}{\mathcal{Z}}
        \frac{k_P^+ + k_P^- + k_O + r + \gamma}
            {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2},
\end{equation}
as stated in Eq.~\ref{eq:model3_fano} in the main text.
